Variable Selection in Convex Function Estimation

### 2014-05-15 tex/scam/additive_faithful.tex
acdc screen

\begin{remark} 
It is difficult to prove the opposite direction of additive faithfulness, that is, if $f$ does not depend on coordinate $k$, then $f_k^*$ will be zero in the additive approximation. Consider as a conceptual example a 3D distribution over $(X_1, X_2, X_3)$; suppose $X_1, X_2$ are independent, and $f$ is only a function of $X_1, X_2$. We can then let $X_3 = f(X_1, X_2) - f^*_1(X_1) - f^*_2(X_2)$, that is, we let $X_3$ exactly capture the additive approximation error, then the best additive approximation of $f$ would have a component $f^*_3(X_3) = X_3$ even though $f$ does not depend on $X_3$. 
\end{remark}

@@

Although convex functions are additively faithful, it is difficult to estimate the optimal additive functions $f^*_k$'s (as defined in equation~\ref{eqn:unconstrained_additive}) because $f^*_k$ need not be a convex function.

Because the true regression function $f$ is convex, it is natural to consider an convex additive model:
\begin{equation}
\label{eqn:convex_additive}
\{ f^*_k \}_{k=1}^s = \arg\min \Big \{ 
    \E\big( f(X) - \sum_{k=1}^s f_k(X_k) \big)^2 \,:\, f_k \in \mathcal{C}^1, \, \E f_k(X_k) = 0 \Big \}
\end{equation}
where $\mathcal{C}^1$ is the set of univariate convex functions.

The convex functions $f^*_k$'s are not additively faithful by themselves but faithfulness can be restored by coupling the $f^*_k$'s with a set of concave functions:
\begin{equation}
\label{eqn:concave_postprocess}
g^*_k = \arg\min \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1, \E g_k(X_k) = 0 
  \Big\}
\end{equation}

>>>>

\begin{remark} 
It remains difficult to find conditions under which the opposite direction of additive faithfulness holds, that is, if $f$ does not depend on coordinate $k$, then $f_k^*$ will be zero in the additive approximation. As a conceptual example, suppose that $f$ is only a function of $X_1, X_2$, and that $(X_1, X_2, X_3)$ follows a degenerate 3-dimensional distribution where $X_3 = f(X_1, X_2) - f^*(X_1) - f^*_2(X_2)$, that is, we let $X_3$ exactly capture the additive approximation error. In this case, the best additive approximation of $f$ would have a component $f^*_3(X_3) = X_3$ even though $f$ does not depend on $X_3$. 
\end{remark}

It is possible to get a similar result for distributions with unbounded support, by using a limit condition $\lim_{|x_k| \rightarrow \infty} \frac{\partial p(\mathbf{x}_{-k} \given x_k)}{\partial x_k} = 0$ instead of the boundary condition. Such a limit condition is too strong however, it is not obeyed by the multivariate Gaussian distribution. \\

In fact, not all convex functions are additively faithful under certain multivariate Gaussian distributions.

\begin{example}
\label{examp:gaussian_counterexample}
Consider a quadratic form $f( \mathbf{x}) = \mathbf{x}^\tran H \mathbf{x} + c^\tran \mathbf{x}$ and a Gaussian distribution $X \sim N(0, \Sigma)$ where $\Sigma_{jj} = 1$ for all $j$.

Then we have a closed form for the additive approximation, which we prove in [TODO:appendix reference].
\begin{itemize}
\item If $f$ does not depend on $x_j$, then $f^*_j(x_j) = 0$.
\item If $f$ depends on $x_j$, then
\[
f^*_j(x_j) = H_j^\tran \Sigma_j x_j^2 + c_j x_j
\]
where $H_j, \Sigma_j$ are the $j$-th row of $H_j, \Sigma_j$ respectively. 
\end{itemize}
Let $H = [1, 2; 2,5]$ and let $\Sigma = [1, -\alpha; -\alpha, 1]$. It is easy to check then that if $\alpha=0.5$, then $f^*_1 = 0$ and additive faithfulness is violated, and if $\alpha > 0.5$, then $f^*_1$ is a concave function.
\end{example}

@@

Although convex functions are additively faithful, it is difficult to estimate the optimal additive functions $f^*_k$'s (as defined in equation~\ref{eqn:unconstrained_additive}) because $f^*_k$ need not be a convex function, as example~\ref{examp:gaussian_counterexample} shows. $f^*_k$ can be shown to be smooth but it is preferable to use shape-constraints whenever possible to avoid the introduction of the additional smoothing bandwidth.

Since the true regression function $f$ is convex, it is natural to ask when it is sufficient to estimate an convex additive model:
\begin{equation}
\label{eqn:convex_additive}
\{ f^*_k \}_{k=1}^s = \arg\min \Big \{ 
    \E\big( f(X) - \sum_{k=1}^s f_k(X_k) \big)^2 \,:\, f_k \in \mathcal{C}^1, \, \E f_k(X_k) = 0 \Big \}
\end{equation}
where $\mathcal{C}^1$ is the set of univariate convex functions.

The convex functions $f^*_k$'s are not additively faithful by themselves unfortunately. But, faithfulness can be restored by coupling the $f^*_k$'s with a set of concave functions:
\begin{equation}
\label{eqn:concave_postprocess}
g^*_k = \arg\min \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1, \E g_k(X_k) = 0 
  \Big\}
\end{equation}

### 2014-05-15 tex/scam/appendix.tex
acdc screen

The Lagrangian is
\begin{equation}
\label{eqn:full_lagrange}
\mathcal{L}( \{ d_k \}, \nu) = 
  \frac{1}{2n} \Big\| 
    Y - \sum_{k=1}^p  \bar{\Delta}_k d_k  \Big\|_2^2 + 
    \lambda \sum_{k=1}^p \| \bar{\Delta}_k d_k \|_\infty -
    \sum_{k=1}^p \sum_{i=1}^{n-1} \nu_{ki} d_{ki} 
\end{equation}
with the constraint that $\nu_{ki} \geq 0$ for all $k,i$.

@@

@@

The Lagrangian of optimization~\ref{opt:alternate_opt_concave} is:
\[
\mathcal{L}_{\trm{cave}}(d_k, \nu_k) = 
  \frac{1}{2n} \big\| \hat{r} - \bar{\Delta}_k d_k \big \|_2^2 + 
  \lambda \| \bar{\Delta}_k d_k \|_\infty + \sum_{k=1}^p \sum_{i=1}^{n-1} \nu_{ki} d_{ki}
\]

>>>>

The Lagrangian is
\begin{equation}
\label{eqn:full_lagrange}
\mathcal{L}( \{ d_k \}, \nu) = 
  \frac{1}{2n} \Big\| 
    Y - \sum_{k=1}^p  \bar{\Delta}_k d_k  \Big\|_2^2 + 
    \lambda \sum_{k=1}^p \| \bar{\Delta}_k d_k \|_\infty -
    \sum_{k=1}^p \sum_{i=2}^{n-1} \nu_{ki} d_{ki} 
\end{equation}
with the constraint that $\nu_{ki} \geq 0$ for all $k,i$.

@@

We have thus proven that there exist one solution $\{ \hat{d}_k \}_{k=1,...,p}$ such that $\hat{d}_k = 0$ for all $k \in S^c$. Furthermore, we have shown that the subgradient variables $\mathbf{u}_k$ of the solution $\{ \hat{d}_k \}$ can be chosen such that $\| \mathbf{u}_k \|_1 < 1$ for all $k \in S^c$.  We now prove that if $\{ \hat{d}'_k \}_{k = 1,..., p}$ is another solution, then it must be that $\hat{d}'_k = 0$ for all $k \in S^c$ as well.  \\

%% prove uniqueness of sparsity pattern.

We first claim that $\sum_{k=1}^p \bar{\Delta}_k \hat{d}_k = \sum_{k=1}^p \bar{\Delta}_k \hat{d}'_k$. If this were not true, then a convex combination of $\hat{d}_k, \hat{d}'_k$ would achieve a strictly lower objective on the quadratic term. More precisely, let $\zeta \in [0,1]$. If $\sum_{k=1}^p \bar{\Delta}_k \hat{d}'_k \neq \sum_{k=1}^p \bar{\Delta}_k \hat{d}_k$, then $\| Y - \sum_{k=1}^p \bar{\Delta}_k \big( \hat{d}_k + \zeta ( \hat{d}'_k - \hat{d}_k) \big) \|_2^2$ is strongly convex as a function of $\nu$. Thus, it cannot be that $\hat{d}_k$ and $\hat{d}'_k$ both achieve optimal objective and we have reached a contradiction.\\

Now, we look at the stationarity condition for both $\{ \hat{d}_k \}$ and $\{ \hat{d}'_k \}$. Let $\mathbf{u}_k \in \partial \| \bar{\Delta}_k \hat{d}_k \|_\infty$ and let $\mathbf{u}'_k \in \partial \| \bar{\Delta}_k \hat{d}'_k \|_\infty$ be the two sets of subgradients. Let $\{ \nu_{ki} \}_{k=1,..,p,\, i=1,...n-1}$ and $\{ \nu'_{ki} \}$ be the two sets of positivity dual variables. \footnote{since there is no positivity constraint on $d_{k1}$, we let $\nu_{k1} = 0$ always.}

Let us define $\bar{\Delta}$, a $n \times p(n-1)$ matrix, to denote the column-wise concatenation of $\{ \bar{\Delta}_k \}_k$ and $\hat{d}$, a $p(n-1)$ dimensional vector, to denote the concatenation of $\{ \hat{d}_k \}_k$. With this notation, we can express $\sum_{k=1}^p \bar{\Delta}_k \hat{d}_k = \bar{\Delta} \hat{d}$.

Since both solutions $(\hat{d}, \mathbf{u}, \nu)$ and $(\hat{d}', \mathbf{u}', \nu')$ must satisfy the stationarity condition, we have that:
\[
\bar{\Delta}^\tran ( Y - \bar{\Delta} \hat{d} ) 
   + \lambda \sum_{k=1}^p \bar{\Delta}_k^\tran \mathbf{u}_k - \nu = 
\bar{\Delta}^\tran ( Y - \bar{\Delta} \hat{d}' ) 
   + \lambda \sum_{k=1}^p \bar{\Delta}_k^\tran \mathbf{u}'_k - \nu' = 0
\] 
We multiply both sides of the above equation by $\hat{d}'$:
\[
\hat{d}'^{\tran}  \bar{\Delta}^\tran ( Y - \bar{\Delta} \hat{d} ) 
    + \lambda \sum_{k=1}^p \hat{d}'^\tran_k \bar{\Delta}_k^\tran \mathbf{u}_k - \hat{d}'^\tran \nu = \hat{d}'^{\tran}  \bar{\Delta}^\tran ( Y - \bar{\Delta} \hat{d}' ) 
    + \lambda \sum_{k=1}^p \hat{d}'^\tran_k \bar{\Delta}_k^\tran \mathbf{u}'_k - \hat{d}'^\tran \nu'
\]
Since $\bar{\Delta} \hat{d}_k = \bar{\Delta} \hat{d}$, $\hat{d}'^\tran \nu' = 0$ (complementary slackness), and $\hat{d}'^\tran_k \bar{\Delta}_k^\tran \mathbf{u}'_k  = \| \hat{f}'_k \|_\infty$ (where $\hat{f}'_k = \bar{\Delta}_k \hat{d}'_k$), we have that:
\[
\lambda \sum_{k=1}^p \hat{d}'^\tran_k \bar{\Delta}_k^\tran \mathbf{u}_k - \hat{d}'^\tran \nu = \lambda \sum_{k=1}^p \| \hat{f}'_k \|_\infty
\]
On one hand, $\hat{d}'$ is a feasible solution so $\hat{d}'^\tran \nu \geq 0$ and so 
\[
\sum_{k=1}^p \hat{d}'^\tran_k \bar{\Delta}_k^\tran \mathbf{u}_k \geq \sum_{k=1}^p \| \hat{f}'_k \|_\infty .
\]

On the other hand, by Holder's inequality:
\begin{align*}
\sum_{k=1}^p \hat{d}'^\tran_k \bar{\Delta}_k^\tran \mathbf{u}_k &\leq 
   \sum_{k=1}^p \| \hat{f}'_k \|_\infty \|\mathbf{u}_k \|_1 
\end{align*}

Since $\mathbf{u}_k$ can be chosen so that $\| \mathbf{u}_k \|_1 < 1$ for all $k \in S^c$, we would get a contradiction if $\| \hat{f}'_k \|_\infty > 0$ for some $k \in S^c$. We thus conclude that $\hat{d}'$ must follow the same sparsity pattern.\\

@@

The Lagrangian of optimization~\ref{opt:alternate_opt_concave} is:
\[
\mathcal{L}_{\trm{cave}}(d_k, \nu_k) = 
  \frac{1}{2n} \big\| \hat{r} - \bar{\Delta}_k d_k \big \|_2^2 + 
  \lambda \| \bar{\Delta}_k d_k \|_\infty + \sum_{k=1}^p \sum_{i=2}^{n-1} \nu_{ki} d_{ki}
\]
